---
title: "Data Manipulation with `dplyr`"
output: github_document
---
```{r, echo = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    fig.width = 6,
    fig.asp = .6,
    out.width = "90%"
)
```

Once you've imported data, you're going to need to do some cleaning up. 

```{r}
library(tidyverse)

options(tibble.print_min = 3)

litters_data = 
  read_csv("./data/FAS_litters.csv")
litters_data =
  janitor::clean_names(litters_data)

pups_data = read_csv("./data/FAS_pups.csv")
pups_data = janitor::clean_names(pups_data)
```

### `select`
Select is useful when you want to select certain columns from your dataset for your analysis.

Ex: selecting 4 variables from dataset `litters_data`: 
```{r}
select(litters_data, group, litter_number, gd0_weight, pups_born_alive)
```
Can specify a range of variables to include:

Ex: select all variables from dataset `litters_data` ranging from `group` variable to `gd_of_birth` variable
To do this, you need to know how your variables are arranged in your dataset
```{r}
select(litters_data, group:gd_of_birth)
```

You can also specify columns you'd like to remove:

Ex: we want everything except for the variable `pups_survive`
```{r}
select(litters_data, -pups_survive)
```

Or we want to remmove multiple variables:
```{r}
select(litters_data, -pups_survive, -group)
```

You can rename variables as part of this process:
Ex: want to select two columns - `group` and `litter_number` and rename them GROUP and LiTtEr_NuMbEr respectively
```{r}
select(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)
```

If all you want to do is rename something, you can use `rename` instead of `select`. This will rename the variables
you care about, and keep everything else: 
```{r}
rename(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)
```

There are some handy helper functions for `select`; read about all of them using `?select_helpers`. I use 
`starts_with()`, `ends_with()`, and `contain()` often, especially when there are variables named with suffixes
or other standard patterns:

```{r}
select(litters_data, starts_with("gd"))
select(litters_data, ends_with("weight"))
```

I also frequently use `everything()`, which is handy for reorganizing columns without discarding anything:

```{r}
select(litters_data, litter_number, pups_survive, everything())
```

### `filter`

You will often filter using comparison operators (`>`, `>=`, `<`, `<=` `==`, and `!=`). You may also use `%in%` to detect if values appear in a set, and `is.na()` to find missing values. The results of comparisons are logical -- the statement is `TRUE` or `FALSE` depending on the values you compare -- and can be combined with other comparisons using the logical operations `&` and `|`, or negated using `!`. 

* `gd_of_birth == 20` 
* `pups_born_alive >- 2`
* `pups_survive != 4`
* `!(pups_survive == 4)`
* `group %in% c ("Con7", "Con8")`
* `group == "Con7" & gd_of_birth == 20`

```{r}
filter(litters_data, gd_of_birth == 20)
```

```{r}
filter(litters_data, group == "Con7" & gd_of_birth == 20)
```

A very common filtering step requires you to omit missing observations. You *can* do this with `filter`, but I recommend using `drop_na` from the `tidyr` package:

* `drop_na(litters_data)` will remove any row with a missing value 
* `drop_na(litters_data, wt_increase)` will remove rows for which `wt_increase` is missing

### `mutate` 

Sometimes you need to select columns; somtimes you need to change them or create new ones. Can do that using `mutate`.

Ex: make a new variable called `wt_gain`, which is the difference between `gd18_weight` and `gd0_weight`

```{r}
mutate(litters_data,
       wt_gain = gd18_weight - gd0_weight,
       group = str_to_lower(group),
       # wt_gain_kg = wt_gain* 2.2
)
```

### `arrange` 

In comparison to the preceding, arranging is pretty straightforward. You can arrange the rows in your data according to the values in on or more columns: 

```{r}
head(arrange(litters_data, group, pups_born_alive), 10)
```

You can also sort in descending order if you'd like.

```{r}
head(arrange(litters_data, desc(group), pups_born_alive), 10)
```

### `%>%`

For example, you want to load data, clean column names, remove `pups_survive`, and create `wt_gain`. You can do a multi-step data manipulation with nest function calls as below: 

```{r}
litters_data_clean = 
  drop_na(
    mutate(
      select(
        janitor::clean_names(
          read_csv("./data/FAS_litters.csv", col_types = "ccddiiii")
          ), 
      -pups_survive
      ),
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)
    ),
  wt_gain
  )
```

But difficult because you have to read it inside out. Piping solves this problem, Allows you to turn the nested approach into a sequential chain by passing the results of one function call as an argument for the next function call. 

```{r}
litters_data = 
  read_csv("./data/FAS_litters.csv", col_types = "ccddiiii") %>%
  janitor::clean_names() %>%
  select(-pups_survive) %>%
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    group = str_to_lower(group)) %>% 
  drop_na(wt_gain)

litters_data
```
The easiest way to read `%>%` is "then". The functions in `dplyr` are designed to work smoothly with the pip operator. By default, the pip will take the result of one function call and use that as the first argument of the next function call; by design, functions in `dplyr` will take a tibble as an input and return a tibble as a result. 


