---
title: "Data Import"
output: github_document
---

```{r, echo = FALSE}
library(tidyverse)
library(readxl)
library(haven)
```

## Data Import: CSVs 

Let's import data using the `readr` package

```{r}
litters_df = read_csv("data/FAS_litters.csv")
litters_df = janitor::clean_names(litters_df)
```

Look at the data 

```{r}
litters_df

head(litters_df)
tail(litters_df)
```

```{r, eval = FALSE}
view(litters_df)
```
Don't include view in R Markdown, will break R Markdown

```{r}
skimr::skim(litters_df)
```

`read_csv` options

Turn NA, 999, and 88 variables into missing variables
Skip the first two rows:
```{r, eval = FALSE}
read_csv("data/FAS_litters.csv", na = c("", "NA", 999, 88), skip = 2)
```

## Other file formats 

We need to read in an excel spreadsheet 

```{r}
mlb_df = read_excel("data/mlb11.xlsx")
```

```{r, eval = FALSE}
view(mlb_df)
```

Specifying range of excel cells in excel sheet when importing excel file: 
```{r}
lotr_words_df = 
  read_excel(
    "data/LotR_Words.xlsx",
    range = "B3:D6")
```

## Still more formats

Read in a SAS dataset 

```{r}
pulse_df = read_sas("data/public_pulse_data.sas7bdat")
```



## Data export

```{r}
write_csv(lotr_words_df, file = "results/lotr_words_df.csv")
```



## Why not base R???
Don't use `read.csv`, always use `read_csv`
```{r}
dont_do_this_df = read.csv("data/FAS_litters.csv")
```




